---
title: "Code Walkthrough for Final Presentation"
author: "Jiahao Zhang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
.libPaths(c("r-lib", .libPaths()))

library(tidyverse)
library(tidymodels)
library(themis)
library(janitor)

set.seed(30100)
```

# 1. Load the cleaned analysis file and define the target

```{r load-and-target, echo=TRUE}
stage3 <- readr::read_csv("Data/stage3.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(
    date = as.Date(date),
    year = as.integer(format(date, "%Y")),
    month = as.integer(format(date, "%m")),
    fatal = factor(if_else(n_killed > 0, "fatal", "non_fatal"),
                   levels = c("non_fatal", "fatal"))
  )
```

This is the main target construction step. I start from the cleaned local file and define the binary outcome `fatal` from `n_killed > 0`.

# 2. Build the final structured feature set

```{r feature-construction, echo=TRUE}
analysis_df <- stage3 |>
  transmute(
    fatal,
    year = factor(year),
    month = factor(month),
    day_of_week = factor(weekdays(date)),
    weekend = factor(if_else(lubridate::wday(date, week_start = 1) >= 6, "weekend", "weekday")),
    state = factor(state),
    city_or_county = factor(city_or_county),
    gun_stolen = factor(gun_stolen),
    gun_type = factor(gun_type),
    n_guns_involved,
    latitude,
    longitude,
    congressional_district = factor(congressional_district),
    state_house_district = factor(state_house_district),
    state_senate_district = factor(state_senate_district),
    addr_has_block = factor(if_else(str_detect(replace_na(address, ""), regex("block", ignore_case = TRUE)), "yes", "no")),
    addr_has_intersection = factor(if_else(str_detect(replace_na(address, ""), regex(" and | & | at ", ignore_case = TRUE)), "yes", "no")),
    addr_has_road = factor(if_else(str_detect(replace_na(address, ""), regex("road|\\brd\\b|street|\\bst\\b|avenue|\\bave\\b|boulevard|\\bblvd\\b", ignore_case = TRUE)), "yes", "no")),
    has_teen = factor(if_else(str_detect(replace_na(participant_age_group, ""), stringr::fixed("Teen 12-17")), "yes", "no")),
    has_child = factor(if_else(str_detect(replace_na(participant_age_group, ""), stringr::fixed("Child 0-11")), "yes", "no"))
  )
```

This is the final feature-engineering step. I turn raw date fields, address text, and age-group text into simpler structured indicators that can go into the model.

# 3. Split the data and build a leakage-safe recipe

```{r split-and-recipe, echo=TRUE}
split_obj <- initial_split(analysis_df, prop = 0.80, strata = fatal)
train_data <- training(split_obj)
test_data <- testing(split_obj)

folds <- vfold_cv(train_data, v = 3, strata = fatal)

basic_recipe <- recipe(fatal ~ ., data = train_data) |>
  step_indicate_na(n_guns_involved, latitude, longitude) |>
  step_impute_median(all_numeric_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_other(city_or_county, gun_stolen, gun_type, threshold = 0.01, other = "other") |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

normalized_recipe <- basic_recipe |>
  step_normalize(all_numeric_predictors())
```

This is the main leakage-safe preprocessing pipeline. All learned steps are estimated on the training data only.

# 4. Define the three comparison models

```{r model-specs, echo=TRUE}
rf_spec <- rand_forest(mtry = 25, min_n = 5, trees = 400) |>
  set_engine("ranger", probability = TRUE) |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = 363,
  tree_depth = 2,
  learn_rate = 0.0187,
  mtry = 8,
  min_n = 13,
  loss_reduction = 2.31,
  sample_size = 0.727
) |>
  set_engine("xgboost", eval_metric = "auc") |>
  set_mode("classification")

logit_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")
```

Here I define the three primary models used in the final comparison: logistic regression, random forest, and XGBoost.

# 5. Show the fitted workflows

```{r model-workflows, echo=TRUE}
rf_fit <- workflow() |>
  add_recipe(basic_recipe) |>
  add_model(rf_spec) |>
  fit(train_data)

xgb_fit <- workflow() |>
  add_recipe(basic_recipe) |>
  add_model(xgb_spec) |>
  fit(train_data)

logit_fit <- workflow() |>
  add_recipe(normalized_recipe) |>
  add_model(logit_spec) |>
  fit(train_data)
```

This is where I combine preprocessing and modeling in a single `workflow()` and fit each model on the training data.

# 6. Tune the tree-based models with cross-validation

```{r tuning-setup, echo=TRUE, eval=FALSE}
roc_auc_fatal <- metric_tweak("roc_auc_fatal", roc_auc, event_level = "second")
metric_fn <- metric_set(roc_auc_fatal, accuracy, bal_accuracy)

folds <- vfold_cv(train_data, v = 3, strata = fatal)

rf_tuned <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_fn,
  control = control_grid(save_workflow = TRUE)
)

xgb_tuned <- tune_grid(
  xgb_workflow,
  resamples = folds,
  grid = xgb_grid,
  metrics = metric_fn,
  control = control_grid(save_workflow = TRUE)
)
```

The tuning step is only for the tree-based models. I use 3-fold cross-validation and select the final tree specifications by ROC AUC for the fatal class.

# 7. Class weighting as one imbalance strategy

```{r class-weighting, echo=TRUE, eval=FALSE}
class_ratio <- sum(train_data$fatal == "non_fatal") / sum(train_data$fatal == "fatal")

rf_weighted_fit <- workflow() |>
  add_recipe(basic_recipe) |>
  add_model(
    rand_forest(mtry = 25, min_n = 5, trees = 400) |>
      set_engine(
        "ranger",
        probability = TRUE,
        class.weights = c(non_fatal = 1, fatal = class_ratio)
      ) |>
      set_mode("classification")
  ) |>
  fit(train_data)
```

This version keeps the same training rows but penalizes mistakes on fatal incidents more heavily.

# 8. Oversampling as another imbalance strategy

```{r oversampling, echo=TRUE, eval=FALSE}
rf_oversampled_recipe <- recipe(fatal ~ ., data = train_data) |>
  step_upsample(fatal, over_ratio = 1, seed = 30100) |>
  step_indicate_na(n_guns_involved, latitude, longitude) |>
  step_impute_median(all_numeric_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_other(city_or_county, gun_stolen, gun_type, threshold = 0.01, other = "other") |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

rf_oversampled_fit <- workflow() |>
  add_recipe(rf_oversampled_recipe) |>
  add_model(rf_spec) |>
  fit(train_data)
```

This version changes the training distribution by resampling the minority class to a 1:1 ratio.

# 9. Threshold moving as the main decision-rule trick

```{r threshold-logic, echo=TRUE}
choose_threshold <- function(fit_obj, val_data) {
  pred <- val_data |>
    bind_cols(predict(fit_obj, val_data, type = "prob"))

  map_dfr(seq(0.05, 0.95, by = 0.01), function(threshold) {
    pred_class <- factor(
      if_else(pred$.pred_fatal >= threshold, "fatal", "non_fatal"),
      levels = c("non_fatal", "fatal")
    )

    tibble(
      threshold = threshold,
      bal_accuracy = bal_accuracy_vec(pred$fatal, pred_class),
      sensitivity = sens_vec(pred$fatal, pred_class, event_level = "second"),
      specificity = spec_vec(pred$fatal, pred_class, event_level = "second")
    )
  }) |>
    arrange(desc(bal_accuracy), abs(threshold - 0.5)) |>
    slice(1)
}
```

This is the main special trick in the project. Instead of stopping at the default cutoff, I search over thresholds and choose the best one on the validation split.

# 10. Add a more conservative specificity-constrained threshold rule

```{r threshold-constrained, echo=TRUE}
choose_threshold_constrained <- function(fit_obj, val_data, min_specificity = 0.75) {
  pred <- val_data |>
    bind_cols(predict(fit_obj, val_data, type = "prob"))

  map_dfr(seq(0.05, 0.95, by = 0.01), function(threshold) {
    pred_class <- factor(
      if_else(pred$.pred_fatal >= threshold, "fatal", "non_fatal"),
      levels = c("non_fatal", "fatal")
    )

    tibble(
      threshold = threshold,
      bal_accuracy = bal_accuracy_vec(pred$fatal, pred_class),
      sensitivity = sens_vec(pred$fatal, pred_class, event_level = "second"),
      specificity = spec_vec(pred$fatal, pred_class, event_level = "second")
    )
  }) |>
    filter(specificity >= min_specificity) |>
    arrange(desc(bal_accuracy), abs(threshold - 0.5)) |>
    slice(1)
}
```

This is the final operating-rule step. I still compare thresholds by balanced accuracy, but only among thresholds that keep specificity near the target bound.

